\chapter{Introduction and Motivation}

\section{Phytoplankton and the Red Sea Ecology: Importance, Impact, Large-Scale
Features, and Applications}

\subsection{The Importance of Phytoplankton}

Phytoplankton are unicellular, free-floating, photosynthetic algae that live in 
the upper layers of bodies of water (ocean, lakes, rivers or ponds). There 
exists a wide diversity of phytoplankton species. Up to date, about 5000 of them 
have been identified \citep{Tett1995}. Phytoplankton are also highly variable in 
sizes, ranging from 0.2μm for cyanobacteria to 200μm for the largest species of 
diatom \citep{Pal2014}. In the oceans, phytoplankton live in the surface layer 
where there is enough sunlight for photosynthesis. 

Phytoplankton plays a fundamental role for the ocean ecology. It is the basis of 
the marine food web and traps most of the energy used by pelagic ecosystems 
\citep{Pal2014}. Zooplankton graze phytoplankton which are then consumed by 
higher trophic levels. It has been estimated that nearly 98\% of the ocean 
primary productivity comes from phytoplankton \citep{Pal2014}. Phytoplankton are 
also responsible for maintaining the dissolved oxygen level necessary for other 
species to survive. However, high phytoplankton concentration may also impact 
their environment by creating dead zones. When they die and sink, the bacteria 
that decompose them can consume all the available oxygen \citep{Pal2014}, 
causing massive mortality in the fauna. Because of its short life cycle, 
phytoplankton respond very well to changes in its environment, making it a key 
parameter to monitor water quality \citep{Wu2014}.

Phytoplankton place at the bottom of the marine food chain makes it an important 
factor for fisheries. Productive fishing zones like the regions in the Arabian 
seas, Californian coast, north-west African coast and Chilean coast are 
explained by the upwelling of cold nutrient rich water favourable to 
phytoplankton growth. On the other hand, the El-Nino phenomenon creates less 
favourable conditions for phytoplankton in the Eastern Pacific, resulting in a 
dramatic reduction of fish catches of fisheries in the western coast of South 
America \citep{Robinson2010}. Remotely-sensed chlorophyll data have been 
routinely used since the last decade to help fisheries predict the timing of 
phytoplankton blooms \citep{Robinson2010}. In contrast, in the Red Sea, the MEI 
(Multivariate ENSO Index) has been found to positively correlate with 
chlorophyll concentration, a fact that could be of importance for regional 
fisheries \citep{Raitsos2015}.

Phytoplankton also plays the role of a biological CO2 pump and strongly impact 
the Earth climate. During photosynthesis, phytoplankton captures carbon and 
releases oxygen. A part of this organic material stays in the food web, either 
transmitted to higher trophic level, or degraded  by bacteria. Another part, 
however, sinks to the bottom of the ocean and forms sediment. It is estimated 
that phytoplankton accounts for 48\% of Earth carbon fixation \citep{Pal2014}.

\subsection{Red Sea Large-Scale Phytoplankton Dynamics}

Typical tropical seas (TTS), like the Red Sea, are characterized by a highly 
stratified structure, where warm nutrient-depleted surface water is separated 
from the cold nutrient-rich deep water by a steep gradient of temperature zone 
called pycnocline. The pycnocline acts as barrier that limits the upward 
nutrients flow \citep{Mann2006}. As a result, TTS are oligotrophic and have low 
chlorophyll concentrations. Until recently, marine biologists believed that 
tropical and subtropical seas have therefore a very low productivity. However, 
recent investigations have contested this idea, suggesting that different 
upwelling mechanisms exist, which bring new nutrients to the surface water 
\citep{Mann2006}.

Despite being an oligotrophic and challenging environment for marine life, the 
Red Sea presents a surprisingly rich and diverse ecosystem \citep{Raitsos2011}, 
and a very well developed coral reef system \citep{Racault}. The source of 
nutrient for sustaining such a developed ecosystem is not well understood yet, 
but the exchange with the open ocean and transport through the mesoscale eddies 
are believed to play an important role \citep{Raitsos2013, Zhan2014}.

Although the Red Sea environment is relatively preserved, it is stressed by 
human activities. The increasing urbanization and fishing activity contribute to 
the fragilization of this unique ecosystem \citep{Acker2008}. An abrupt increase 
of temperature has further occurred in the last decade which may threaten the 
fragile coral reef system \citep{Raitsos2011}.

Because of the lack of in-situ data, the large-scale phytoplankton dynamics of 
the Red Sea remain largely unknown \citep{Raitsos2013, Triantafyllou2014}. 
However, in recent studies, remotely-sensed data and computer simulations have 
been used to improve our knowledge of the ecology of this region. The Red Sea is 
deficient in major nutrients \citep{Weikert1987}, and the only significant input 
of water comes from the Gulf of Aden. This explains a general pattern of 
chlorophyll concentration increase from north to south \citep{Raitsos2013}, with 
the lowest concentration found in the northern central Red Sea. The Red Sea also 
displays a distinct seasonality, with a peak in concentration during the winter. 
A weak summer peak is also observed around July, everywhere except in the 
northernmost region \citep{Raitsos2013}. Despite this regularity, a strong 
interannual variability is observed, with blooms that can reach mesotrophic 
concentration levels \citep{Raitsos2013}. According to 
\citet{Triantafyllou2014}, the variations in the Red Sea ecology are mainly 
driven by physical circulation. In the rest of this section, we explore some of 
the mechanisms that have been linked to the major features of chlorophyll 
concentration.

The exchange of water with the nutrient-rich Gulf of Aden is a major driving 
mechanism for the whole Red Sea \citep{Triantafyllou2014}. It is the most 
important source of nutrient. The maximum chlorophyll concentration observed in 
the southern Red Sea during winter is attributed to wind-driven water intrusion 
\citep{Raitsos2013}. In Summer, this exchange of water is believed to be the 
only significant source of nutrients for the whole Red Sea. The influence of the 
water intrusion weakens as the latitude increases, explaining the low 
concentration in the northern half of the Red Sea \citep{Raitsos2013}.

Deep convection also plays an important role in allowing nutrient-rich deep 
water to mix with water of the euphotic zone. The vertical mixing is the most 
vigorous in the northern extremity of the Red Sea during the winter. This 
explains its higher chlorophyll concentration compared to the north-central Red 
Sea, a region of weak mixing \citep{Raitsos2013}. The northern Red Sea mixing is 
believed to be driven by wind \citep{Raitsos2013}.

The Red Sea circulation is strongly influenced by mesoscale eddies 
\citep{Yao2014, Zhan2014} that could impact primary production \citep{Zhai2013}. 
In particular, the anti-cyclonic eddy in the central Red Sea is believed to 
control the June concentration peak and the summer productivity of this region, 
by transporting nutrients and/or phytoplankton from the adjacent coral reefs 
\citep{Raitsos2013}. In the northern Red Sea, a cold-core eddy plays a role in 
enhancing the vertical mixing in that region \citep{Raitsos2013}.

Aerial depositions of dust could also be an important input of nutrient for the 
Red Sea, but it has been largely left unexplored \citep{Triantafyllou2014}. 
\citet{Raitsos2013} noticed for example that sand storms in the Red Sea most 
frequently happen in June and July, which coincides with the summer chlorophyll 
peak. Finally, climate mode indices have been shown to be strongly correlated 
with air-sea heat exchanges in the Red Sea \citep{Abulnaja2015}, and might 
therefore influence its biology.

\section{Remotely-Sensed Chlorophyll Data: Relevance and Challenges for the Red
Sea}

\subsection{Measuring Chlorophyll Concentration}

Chlorophyll is a molecule present in algae, phytoplankton and plants that is 
critical for photosynthesis. It is a poor absorber of green light, and is 
responsible for the coloration of plants. When phytoplankton are present in high 
concentrations, the water also takes a detectable green coloration (it can also 
take a red or blue coloration depending on the type of dominating phytoplankton) 
\citep{Robinson2010}. This offers a way to estimate the phytoplankton 
concentration.

In-situ measurement of chlorophyll concentration can be gathered through 
scientific cruises, buoy stations or gliders (unmanned submarines). These 
methods are expensive to deploy and therefore generally have limited temporal 
and spatial coverage \citep{Robinson2010}. Political issues, as in the Red Sea, 
as well as security issues as in the Arabian Sea, set also barriers to in-situ 
measurements.

Satellite measurements of chlorophyll provide excellent proxies for 
phytoplankton concentrations with a good temporal and spatial coverage 
\citep{Robinson2010}. The SeaWIFS, MODIS and MERIS missions have provided an 
uninterrupted coverage of the world since 1997. High-resolution maps of daily 
chlorophyll concentration are freely accessible to the scientific community 
\citep{McClain2009}. Despite some limitations, like missing data due to cloud 
coverage and sunglint, or problematic values in coastal areas, remotely-sensed 
chlorophyll concentration are intensively used by the scientific community. In 
regions, like in the Red Sea, where little in-situ measurements are available 
\citep{Raitsos2013, Brewin2013}, these constitute the most important data 
source.

\subsection{Limitation of Remotely-Sensed Chlorophyll Data in the Red Sea}

The quality of remotely-sensed chlorophyll data products such as MODIS and 
SeaWiFS in the Red Sea is comparable with that of the rest of the world for case 
I waters (open sea) \citep{Brewin2013}. However, the data contains a large 
amount of missing values because of persistent clouds, sun-glint and sensor 
saturation \citep{Racault}. This problem is particularly acute during the summer 
in the southern Red Sea where the data coverage is 0\% \citep{Racault}.

Chlorophyll concentration estimation in optically complex case II waters is a 
recurrent problem in this data that particularly affects the southern Red Sea. 
In this region, the remotely sensed chlorophyll data could be overestimated 
\citep{Raitsos2013}. However, all high values are not necessarily bad, as highly 
productive coral reefs are also present in this region \citep{Raitsos2013}. 
However, these values have not been validated yet, due to the lack of in situ 
measurements \citep{Raitsos2013}.

One solution to missing and bad values is to use a data filling algorithm, of 
which one of the most popular is the Date INterpolating Empirical Orthogonal 
Functions (DINEOF). It is an EOF based data filling approach introduced by 
\citet{Beckers2003}. It has been used for multivariate reconstruction of SST 
fields using chlorophyll data in \citet{Alvera2007}. In \citet{Sicarjobs2011}, 
it has been employed to fill chlorophyll data with 70\% of missing values. 
\citet{Taylor2013} has compared DINEOF with other EOF-based reconstruction 
algorithms, suggesting that the former is the best method for data filling. 
DINEOF has been employed in several other chlorophyll studies \citep{Miles2010, 
Waite2013}.

The OC-CCI is a new chlorophyll data product that considerably increases the Red 
Sea coverage. It merges the data from sensors SeaWiFS, MODIS and MERIS. Overall, 
it achieves a 75-80\% coverage in the entire Red Sea basin against 50-65\% for a 
single sensor \citep{Racault}. This is mostly due to the use of the POLYMER 
algorithm \citep{Steinmetz2011} that allows to exploit MERIS data collected 
during hazy conditions. However, this new dataset has not been fully explored to 
revisit the assumptions made on the large-scale Red Sea phytoplankton 
productivity.

\section{Modeling and Forecasting Chlorophyll: Data-Driven and Physics-Driven 
Approaches, and Applications}

\subsection{Why Modeling Chlorophyll?}

Models could be useful to identify causes behind the chlorophyll patterns we 
observe in the Red Sea. Many hypotheses have been made about the drivers of 
chlorophyll concentration in this regions, but some of them have not been yet 
investigated through models. The role played by the exchange of water with the 
Gulf of Aden and winter overturning in the northern Red Sea have been 
successfully modeled with a 3D coupled ecological model 
\citep{Triantafyllou2014}. However, the interaction between the open sea and 
coral reefs, and the role of atmospheric depositions have not been investigated 
yet. Models, can also be helpful for understanding governing dynamics affecting 
the chlorophyll concentration. In particular, the interaction between the 
productivity level of the different regions of the Red Sea is yet to be 
explored.

Model predictions for chlorophyll concentration also have practical 
applications. Phytoplankton blooms can be harmful to humans and marine life and 
are closely monitored in many regions of the world \citep{Pettersson2013}. In 
the Red Sea, where tourism and aquaculture are developing, it is likely to 
become a concern too. Phytoplankton is also directly, and indirectly through 
zooplankton, the cause of microfouling that affects desalination plants. In 
2008-2009, a red tide forced the shutdown of desalination plants along the Gulf 
of Oman and the Persian Gulf \citep{Richlen2010}.


\subsection{Deterministic Models}

\subsubsection{Ecological Models}

Ecological ordinary differential equation (ODE) deterministic models are a 
popular approaches to model marine ecology. These models represent the ecosytem 
as the interactions of complex physical, chemical and biological processes. The 
interactions are modeled by differential equations that represent the flow of 
carbon, nitrogen, phosphate and silicon. The biota is divided into trophic 
levels, and can be further divided by feeding methods and size classes.

These models can be as simple as the nutrient-phytoplankton-zooplankton (NPZ) 
model that only has three variables representing two trophic levels and nitrate, 
or as complex as the European regional seas ecosystem model (ERSEM) that has 
dozens of variables. Such a model has been coupled to the MITgcm circulation 
model used to simulate the Red Sea ecology \citep{Triantafyllou2014}. However 
the complexity of these models makes them difficult to parametrize if not enough 
data are available, which is usually the case \citep{Anderson2005}.

\subsubsection{Data Assimilation}

Data assimilation is used to improve the simulations of ecological dynamics 
models and enhance their forecasting capabilities by correcting their 
predictions with observations. Such prediction capabilities are deployed in 
operational expert systems, for example to study the impact of human activities 
on the ecosystem of the Gulf of Pagasitikos \citep{Korres2012}. The deployment 
of a similar forecasting system in the Red Sea is currently under development 
\citep{Triantafyllou2014}. Hindcasting, the estimation of unobserved variables, 
is another application of assimilation scheme. \citet{Ciavatta2011}  showed that 
they could improve the seasonal and annual hindcast of non-assimilated 
biogeochemical properties in the shelf area of Western English Channel. Data 
assimilation can also be used for reanalysis, to provide estimates of past years 
biogeochemical variables \citep{Fontana2013}. 

In the marine ecology modeling community, three assimilation schemes have been 
widely used: the Ensemble Kalman filters (EnKF), the Singular Evolutive Extended 
Kalman filter (SEEK), and its ensemble variant, the Singular Evolutive 
Interpolated Kalman filter (SEIK). The Stochastic EnKF, a Monte-Carlo 
approximation of the Kalman Filter, has been used in \citet{Ciavatta2011, 
Ciavatta2014}. However, it suffers from sampling errors when the ensemble size 
is smaller than the number of observations, as is usually the case when 
assimilating remotely-sensed data. SEEK is a reduced-rank variant of the 
Extended Kalman filter (EK), which is intractable in high-dimensions. It is 
based  on the projection of the error covariance onto a low dimensional space. 
SEEK has a long history in data assimilation for marine ecology models and is 
still used in recent studies \citep{Fontana2013, Korres2012, Butenschon2012}. 
SEIK is a deterministic version of the EnKF that do not suffer from observations 
sampling errors, as it updates the filter and forecast exactly as in the Kalman 
filter, but requires a resampling step to generate a new ensemble for the next 
forecast step. SEIK has been used by \citep{Triantafyllou2012, Korres2012}. 
\citet{Korres2012} shows that SEIK and SEEK are both comparably robust methods 
for highly nonlinear systems.

Ecological models are challenging applications for data assimilation schemes. 
First, biogeochemical variable are usually positive concentration, whereas 
Kalman filters expect Gaussian variables, and log-transformation can fail at 
solving this issue \citep{Ciavatta2011}. In an attempt to mitigate this problem, 
\citet{Fontana2013} has introduced Gaussian anamorphosis transformations. 
Second, ecological blooms are intermittent and highly nonlinear, conditions that 
are challenging for Kalman filter-based assimilation schemes 
\citep{Triantafyllou2012, Korres2012}. Third, SEIK, EnKF and SEEK project the 
error covariance onto some subspace, resulting in an underestimation of the 
estimation error. \citet{Butenschon2012} studied different ways to propagate the 
error covariance in order to alleviate this issue. Finally, the model error 
statistics are required by Kalman-derived filters, but are difficult to 
estimate. \citet{Triantafyllou2012} proposed to use the $H_\infty$ method with 
SEIK in order alleviate this requirement.

\subsection{Data-Driven Approaches}

Compared to data assimilated ecological models, data-driven statistical models 
are relatively easier to apply. They are relevant when the phenomenon producing 
the data is dynamically complex to model, or simply poorly understood 
\cite{Gareth2013}. They have been applied to predict chlorophyll concentration, 
mostly in small regions that have complex dynamics (see examples below). Some 
statistical models, such as linear regression, Gaussian additive models, or tree 
regression have the advantage of being easy to interpret \citep{Gareth2013}, and 
could be used to understand the dynamics driving the chlorophyll concentration.

\subsubsection{Statistical Models}

Statistical and machine learning models have been used for estimation and 
classification problems related to phytoplankton concentrations. One application 
is the detection of harmful algal bloom from spatio-temporal satellite dataset, 
that has been addressed by \citet{Gokaraju2011} in the Gulf of Mexico using 
support vector machines. Another application is the estimation of chlorophyll 
concentration in case II coastal water using satellite radiance data. This 
problem has been considered by \citet{Kim2014} on the west coast of South Korea, 
and by \citet{Camps-Valls2006} using a global dataset of in situ measurements. 
The former used the support vector regression algorithm, while the latter used 
the random forest algorithm.

Machine learning algorithms, in particular Artificial Neural Networks, have been 
very popular for forecasting regional chlorophyll concentration in regions with 
very complex dynamics. In such regions, deterministic ecological models are 
usually too complex and less efficient than data-driven approaches. Neural 
networks have been widely used for forecasting chlorophyll concentration in 
fresh as well as in coastal water systems. In \citet{Jeong2006}, temporal 
recurrent recursive neural network have been used and found superior to 
traditional time-series model for daily forecasts of chlorophyll concentration. 
\citet{Wang2013} also used recurrent neural networks for daily chlorophyll 
forecasting in Lake Taihu, China. \citet{Mulia2013} combined Neural Network and 
genetic algorithm for nowcasting and forecasting of the chlorophyll 
concentration up to 14 days ahead, in the tidal dominated coast of Singapore. 
Finally, \citet{Lee2013} used neural networks for the forecasting of algal bloom 
with one or two weeks lags in the coastal waters of Hong-Kong.

\subsubsection{Geostatistics}

Phenomena such as propagation and diffusion play a key role in the chlorophyll 
spatial concentration, but are difficult to represent without spatial modeling. 
There is also a difference in the chlorophyll patterns of different regions of 
the Red Sea, in particular between the nutrient rich southern Red Sea and the 
oligotrophic northern Red Sea, and between the open ocean and the coastal waters 
\citep{Raitsos2013}. It is however difficult to come up with a clear division of 
the Red Sea in eco-regions. One should also expect the different regions of the 
Red Sea to interact.

Classical geostatistics is the most widely used spatial statistical model. It 
models spatial data as the realization of a two-dimensional Gaussian process, of 
which one can estimate the parameters. Geostatistics can be easily extended to 
spatio-temporal datasets. Many flexible ways of constructing space-time 
covariance functions for these models have been proposed recently 
\citep{Gneiting2002, Cressie1999, Stein2005}. Space-time geostatistics has been 
applied in many environmental studies, but not to chlorophyll data yet.

The theory of space-time geostatistics is closely related to that of spatial 
statistics. In fact, the time dimension is an additional dimension. However, the 
relationship between these two dimensions derive from a dynamical process, that 
must be taken into account in the definition of the covariance funciton 
\citep{Gneiting2010}. Some space-time covariance models can actually be derived 
from a physical formulation, such as the frozen fields \citep{Gneiting2010}, or 
SDEs \citep{Brown2000, North2011}.

Despite their theoretical interest, physically-derived space-time covariance 
function have been little used \citep{Gneiting2010}. More popular, are 
covariance functions built from simple building blocks. One of the most simple 
types are separable covariance functions, that are the product of a spatial 
covariance function and temporal covariance function. They are computationally 
efficient, but are not suitable to represent space-time interactions 
\citep{Cressie1999, Stein2005}, making them of limited use for modeling physical 
systems. The Cressie, Huang spectral characterization theorem of space-time 
covariance functions has opened the door to wider ways of constructing them. For 
example, \citet{Gneiting2002} presented a simple criterion that allows their 
construction from a very large class of models. 

Space-time geostatistical models have been used in a variety of applications. 
\citet{Hohn1993} used it for forecasting the outbreaks of an invasive specie. 
These methods have also been used in meteorology to model temperature fields 
\citep{Handcock1994, North2011} or wind \citep{Cressie1999, Gneiting2002}, and 
in environmental studies for ground-level ozone concentration. 
\citet{Gneiting2007, Gneiting2010} present recent more details on the theory of 
space-time geostatistics and its applications.

\section{Thesis Objectives}

The goal of this thesis is to develop novel models for chlorophyll forecasting
in the Red Sea. Two approaches will be considered: a data-driven driven
approach that derives a model from data using statistical models, and a
dynamical approach that derives a model from first principles. The merits of both
approaches will be compared for the first time in the same region using the same
data. We will also propose ways to combine both methods for improving chlorophyll
forecasting.

There is a crucial need for improving the modeling of large-scale features
of marine ecosystems. The ocean physical, chemical and ecological processes are
complex and poorly understood. Even though ocean color remote-sensing data has
revolutionized our understanding of marine ecology, it only gives information
about the lower trophic levels at the sea surface. Our knowledge is even more
limited in the Red Sea due to the paucity of in situ data.

Currently, both deterministic and statistical approaches have been used to predict
chlorophyll. However, these approaches have never been compared in the same
problem. A thorough comparison is very much needed. It might help modelers to
have a foundation on which to base their modeling strategy.

This thesis will also help increase our understanding of the Red Sea ecology.
In particular, it will identify possibles drivers for the chlorophyll
seasonality and interannual variability. We will also identify and
characterize the different Red Sea eco-regions. This work has practical
applications for the region. In particular, better forecasting phytoplankton
blooms may be extremely helpful of the operation of desalination plants.
