\section{Chapter 4: Local Geostatistical Model for Chlorophyll Forecasting}

This chapter will bring together the results of the two preceding chapters to
develop a geostatistical model that takes into account the large-scale dynamics
and the regional spatio-temporal dynamics. In chapter 2, a predictive model is
built, that represents the large scales behaviour of the Red Sea, but the
spatial dimension inside each region is not considered. We expect local
features to play an important role, such as the proximity to the coast, the
bathymetry, proximity to other regions or major cities, etc. In task 3, we
proposed a methodology to use a geostatistical model in a dynamical fashion to
do pixel-scale forecast. In this task, each regions will be modeled separately
by a local geostatistical model that can provide local prediction. These models
will have access to aggregate covariates from neighboring regions to represent
the global scale behaviours. 

\paragraph{Open Questions}

\begin{itemize}

\item What are the most adapted space-time covariance models for chlorophyll
data?

\item How to use global covariates in a geostatistical model?

\item What are the differences in the spatio-temporal correlation of
chlorophyll in each region?

\item Can the regional behaviour of phytoplankton be accurately predicted?

\item What are the spatial features that are important for regional chlorophyll
dynamics in the Red Sea?

\end{itemize}

\paragraph{Methods}

\begin{itemize}

\item Extract local datasets from previous tasks based on which we will train
regional geostatistical models.

\item Design the training and test datasets, and the cross-validation method,
as in chapter 2.

\item Estimate the chlorophyll mean functions for each region as function of
the past covariates, to improve the model forecasting skills.

\item Fit local geostatistical models to the residuals, to model them as
space-time Gaussian processes.

\item Evaluate the model predictions and compare the results with chapters 2
and 3.

\end{itemize}

\paragraph{Expected Outcomes}

\begin{itemize}

\item A methodology to aggregate local geostatistical models.

\item Improved prediction skills compared to the models developed in Chapters 2
and 3.

\item An understanding of the differences between the regions that have been
identified in Chapter 2.

\item A critical evaluation of the space-time covariance models for fitting
chlorophyll data.

\item A better characterization of the regional chlorophyll dynamics.

\end{itemize}
