\section{Chapter 6: Combining Statistical and Data Assimilative Predictive
Models for Improved Chlorophyll Forecasting}

In the previous chapters, we proposed to evaluate the forecasts of the
ecological ERSEM model against those of the statistical models we develop in
chapters 2 to 4. In this chapter we will study how these two approaches can be
complementary. Specifically, we will study the use of statistical forecasts
model to improve the forecasts of the ERSEM ecological model. The forecasts of
the statistical models will be treated as observations, that can be assimilated
by filtering or smoothing schemes implemented with the ERSEM model. This will
hopefully lead to improved forecasts. Real observations will be assimilated
sequentially when they become available. This method will allow the different
ERSEM models implemented on each cluster to indirectly communicate their
estimates between each other.

\paragraph{Open Questions}

\begin{itemize}

\item Can statistical predictive models be used to communicate information
between deterministic models?

\item Would the access to information about other regions improve the model
forecasts?

\item What are the global patterns of ecological dynamics in the Red Sea?

\end{itemize}

\paragraph{Methods}

\begin{enumerate}

\item Develop a new assimilation scheme to assimilate statistical predictions
and improve the 1D regional deterministic models.

\item Define metrics to measure model improvement compared the results of Chapter
5.

\item Prepare training and test datasets, as in previous chapters.

\item Train statistical models to predict chlorophyll as in Chapter 2.

\item Run simulations with assimilation of statistical and real observations.

\item Compare with the results of Chapter 5.

\end{enumerate}

\paragraph{Expected Outcomes}

\begin{itemize}

\item An improvement in the prediction skills of the dynamically-driven
approach.

\item A methodology to couple assimilative dynamically-driven regional
ecological models through statistical models.

\item Insights on the global ecological dynamics of the Red Sea.

\item An efficient general approach to model and forecast chlorophyll
concentrations.

\end{itemize}
