\section{Chapter 2: Forecasting Chlorophyll Concentration in Regional Aggregates}

\noindent
\emph{Duration: 2 months (by February 2015)}

\noindent
\emph{Submission: Progress in Oceanography}

\noindent
\emph{Collaborator: Dionysios Raitsos}

Chlorophyll data is very complex. It is therefore useful to first simplify it by aggregating it spatially. The space-time dynamics of the chlorophyll data reflects the highly nonlinear dynamics of the underlying physical, chemical and biological phenomenons. As shown by the north-south gradient and the seasonal behavior, the resulting space-time process is nonstationary in time and in space. The high-dimensionality in space can be reduced by considering a regional aggregation of the results. This would allow us to focus on the global scale phenomenons: such as the interactions between neighboring regions, the time-scale of large events and the difference in the physical variables affecting the chlorophyll concentration in each region. In the following tasks, these simple predictive models will also be a reference for evaluating more complex ones. 

\paragraph{Open Questions}

\begin{itemize}
  \item Is the biological aggregation of the Red Sea proposed in \cite{Raitsos2013} statistically meaningful?
  \item Can clustering methods be used to identify marine ecological zones based on chlorophyll data?
  \item Can a simple forecasting model allow us to understand the causes of chlorophyll blooms?
  \item Can the current hypothesizes about the seasonal chlorophyll dynamics be validated?
\end{itemize}

\paragraph{Method}

\begin{enumerate}
  \item Define datasets (training and test datasets, cross-validation)
  \item Variable selection (Lasso, L1 regression, single-variable linear regression)
  \item Define regional aggregations (unsupervised learning, Hierarchical clustering, K-means)
  \item Forecasts chlorophyll concentration (linear regression, GAM models, diagnostic, k-nearest neighbors)
  \item Predicting future extreme blooms (nearest-neighbours, logistic regression, decision trees)
\end{enumerate}

\paragraph{Expected Outcomes}

\begin{itemize}
  \item A regional division of the Red Sea that has been quantitatively evaluated.
  \item A critical evaluation of current hypothesis about the chlorophyll dynamics in the Red Sea.
  \item A lower bound on the performance of a more sophisticated model.
  \item An assessment of the limitation of aggregate methods for Chlorophyll data.
  \item An understanding on how the treatment of spatial correlations can improve the results.
\end{itemize}
